Field-informed Reinforcement Learning for Learning Large Scale Collective Tasks

### 2023-05-05 paper.tex
chore: add motivation, background draft done

In this paper, we consider a new reinforcement learning approach 
  for the definition of intelligent agents called \emph{Field-Informed reinforcement learning}, 
  where we use computational \emph{computational fields} to manage the interaction between agents 
  in stigmergically and Graph Neural Networks to learn a local behaviour necessary to solve collective tasks. 
%
This allows us to create distributed controllers informed by a collective knowledge 
 distilled during learning but that use only local information at runtime.
% 
We demonstrate the effectiveness of this new approach in several case studies 
 where coordination tasks are successfully solved. \lukas{do we really have multiple case studies?} 
\end{abstract}

@@

One interesting application of \ac{RL} is when there are multiple learning agents involved. 
 Such scenarios are referred to as multi-agent reinforcement learning (MARL)~\cite{zhang2019marl}. 
% While standard \ac{RL} approaches use Markov decision processes to model the world, 
% in MARL for distributed systems, the notion of 
% \acp{DecPOMDP}~\cite{Decpomdp2000} is used.
%
%A \ac{DecPOMDP} represents a sequential game, 
% in which each agent receives a local observation about the world they 
% inhabit and receives a joint immediate reward in response to the actions they take. 
% 
%\ga{Improve, if necessary, the following paragraph.}
%Formally, a \ac{DecPOMDP} is defined as a tuple $\langle N, S, \{A_i\}, P, R, \{\Omega_i\}, O, \gamma \rangle$, 
% where $N$ is the total number of agents that live in the environment, 
% $S$ is the state set of the environment, 
% $\{A_i\}$ is the set of actions for agent $i$ and $A$ is the joint action space, 
% $P$ is the probability transition function, 
% $R$ is the immediate joint reward function, and 
% $\{\Omega_i\}$ is the set observation function for the agent $i$,
% $O$ is the observation space, and finally $\gamma$ is the discount factor.
%
%On top of this model, each agent tries to learn a policy $\pi_i$ in order 
% to maximise the long-term joint reward signal.
%In the case of some of these concepts, 
% there are several approaches/classifications in the literature, 
% like cooperative (i.e., the reward function incentivizes collective outcomes)
% heterogeneous (i.e., the agents have different action spaces and policies) and
% homogenous (i.e., the agents have the same action space and policy) settings.
%  
In particular, in this work, we consider homogenous many agent RL~\cite{yang2021many}, 
 where $N \gg 2$ and each agent is interchangeable and indistinguishable.
%
This research area is relevant in the context of large-scale systems 
 when the incentive of the individual is reduced compared to the collective behaviour of the system, 
 which is found in collective adaptive systems such as swarm robotics.

@@

This model can be used then to express the evolution of the system in time.
Specifically, starting from a global state $\mathcal{S}^P_t$, the next state $\mathcal{S}^P_{t+1}$ is defined as:
\begin{equation}
\mathcal{A}^P_t = \pi(\xi(S^P_t))
\end{equation}
\begin{equation}
\mathcal{S}^P_{t+1} = \mathcal{T}(\mathcal{S}^P_t, \mathcal{A}^P_t)
\end{equation}
Given a time $t$, the system can be also represented as a graph $G_t = (V_t, E_t)$, 
  where $E_t$ is built from $\mathcal{N}$.
  Each node is then decorated with the local observation $o_v$. 
  Therefore, this graph can be used for being processed by a \ac{GNN} as done in~\cite{DBLP:conf/corl/TolstayaGPP0R19,tolstaya2020learning,DBLP:conf/icra/GosrichMLPYR022}.
\section{Field-informed Reinforcement Learning}
\label{sec:approach}
\ga{Page budget: 2/3 pages \\}
\ga{Plan: we should discuss the approach in detail, including the system model, and the learning dynamics.
Particularly, I would like to highlight how each node in the graph is a local controller,
but it could be seen as a global evolution, therefore we can use global information to inform the local controller.}
\subsection{System model}
\ga{Plan: 
  discussion about the typical AC evaluation model, specifying how GNN (with 1-hop information diffusion)
  can be easily integrated with this distributed model
}
\subsection{Learning dynamics}
\ga{
  Plan: discuss the chosen approach, that is Centralised Training and Distributed Execution.
  Particularly, we use DQN in which the neural network is a GNN. So, the dataset consists in a set of graphs,
  where each graph is a snapshot of the system. 
  The target is the Q-value of the action taken in the current state.
  Explain how this can be then used for distributed controllers.
}
\section{Evaluation}
\label{sec:eval}
\ga{Page budget: 2 pages \\}
\ga{Plan: we should discuss the robot aggregation scenario, specifying the state space, the action space, and the reward function.
I dunno if it makes sense to discuss the variants (i.e., fixed area, two areas and moving area). We will see.
}
\subsection{Scenario}

>>>>

In this paper, we consider a new reinforcement learning approach 
  for the definition of intelligent agents called \emph{Field-Informed reinforcement learning}, 
  where we use computational \emph{computational fields} to manage the interaction between agents 
  and Graph Neural Networks to learn a local behaviour necessary to solve collective tasks. 
%
This allows us to create distributed controllers informed by a collective knowledge 
 distilled during learning but that use only local information at runtime.
% 
We demonstrate the effectiveness of this new approach in a simulated use case 
 where coordination tasks are successfully solved. \lukas{do we really have multiple case studies?} 
\end{abstract}

@@

One interesting application of \ac{RL} is when there are multiple learning agents involved. 
 Such scenarios are referred to as multi-agent reinforcement learning (MARL)~\cite{zhang2019marl}. 
% While standard \ac{RL} approaches use Markov decision processes to model the world, 
% in MARL for distributed systems, the notion of 
% \acp{DecPOMDP}~\cite{Decpomdp2000} is used.
%
%A \ac{DecPOMDP} represents a sequential game, 
% in which each agent receives a local observation about the world they 
% inhabit and receives a joint immediate reward in response to the actions they take. 
% 
%\ga{Improve, if necessary, the following paragraph.}
%Formally, a \ac{DecPOMDP} is defined as a tuple $\langle N, S, \{A_i\}, P, R, \{\Omega_i\}, O, \gamma \rangle$, 
% where $N$ is the total number of agents that live in the environment, 
% $S$ is the state set of the environment, 
% $\{A_i\}$ is the set of actions for agent $i$ and $A$ is the joint action space, 
% $P$ is the probability transition function, 
% $R$ is the immediate joint reward function, and 
% $\{\Omega_i\}$ is the set observation function for the agent $i$,
% $O$ is the observation space, and finally $\gamma$ is the discount factor.
%
%On top of this model, each agent tries to learn a policy $\pi_i$ in order 
% to maximise the long-term joint reward signal.
%In the case of some of these concepts, 
% there are several approaches/classifications in the literature, 
% like cooperative (i.e., the reward function incentivizes collective outcomes)
% heterogeneous (i.e., the agents have different action spaces and policies) and
% homogenous (i.e., the agents have the same action space and policy) settings.
%  
In particular, in this work, we consider homogenous \ac{MAARL}~\cite{yang2021many}, 
 where $N \gg 2$ and each agent is interchangeable and indistinguishable.
%
This research area is relevant in the context of large-scale systems 
 when the incentive of the individual is reduced compared to the collective behaviour of the system, 
 which is found in collective adaptive systems such as swarm robotics.

@@

This model can be used then to express the evolution of the system in time.
Specifically, starting from a global state $\mathcal{S}^P_t$, the next state $\mathcal{S}^P_{t+1}$ is defined as:
\begin{equation}
\mathcal{A}^P_t = \pi(\xi(S^P_t))
\end{equation}
\begin{equation}
\mathcal{S}^P_{t+1} = \mathcal{T}(\mathcal{S}^P_t, \mathcal{A}^P_t)
\end{equation}
Given a time $t$, the system can be also represented as a graph $G_t = (V_t, E_t)$, 
  where $E_t$ is built from $\mathcal{N}$.
  Each node is then decorated with the local observation $o_v$. 
  Therefore, this graph can be used for being processed by a \ac{GNN} as done in~\cite{DBLP:conf/corl/TolstayaGPP0R19,tolstaya2020learning,DBLP:conf/icra/GosrichMLPYR022}.
\subsection{Motivation}
The definition of distributed controllers for intelligent systems can be divided into two major approaches: 
 \emph{manual design}, where the application is conceived and described through a programming language, 
 and \emph{automatic design}~\cite{brambilla2013swarm}, which employs machine learning techniques to synthesize distributed controllers. 
 Our work lies between these two approaches, particularly in the areas of field-based coordination and \ac{MAARL} with the use of \acp{GNN}.

Compared to earlier works, 
 our approach builds on the concept of co-fields, 
 where agents exploit the field as a stigmergic medium to receive system-wide information and apply reasoning to this data.
%  
We present a subsequent approach where agent intelligence is synthesized through \ac{MAARL} and the use of \ac{GNN} to learn a local representation from the neighbourhood.
%
By learning a smart policy directly in the environment, the agent becomes capable of adapting its behaviour to new situations.

The use of \acp{GNN} as part of a distributed controller has been explored in previous literature, 
 where it was shown that they could be used to break down the evaluation of local and distributed systems. 
%
However, in these works, communication was left entirely to the neural network, 
 making the learning process potentially more complex and unstable. 
 In our approach, the \ac{GNN} is \emph{informed} by computational fields that collect the necessary information to compute a certain task, 
 limiting learning to only the specific task defined by a collective reward function.
 This will speed up the learning process and make it more stable.
\section{Field-informed Reinforcement Learning}
\label{sec:approach}
\ga{Page budget: 2/3 pages \\}
\ga{Plan: we should discuss the approach in detail, including the system model, and the learning dynamics.
Particularly, I would like to highlight how each node in the graph is a local controller,
but it could be seen as a global evolution, therefore we can use global information to inform the local controller.}
\subsection{System model}
\ga{Plan: 
  discussion about the typical AC evaluation model, specifying how GNN (with 1-hop information diffusion)
  can be easily integrated with this distributed model
}
\subsection{Learning dynamics}
\ga{
  Plan: discuss the chosen approach, that is Centralised Training and Distributed Execution.
  Particularly, we use DQN in which the neural network is a GNN. So, the dataset consists in a set of graphs,
  where each graph is a snapshot of the system. 
  The target is the Q-value of the action taken in the current state.
  Explain how this can be then used for distributed controllers.
}
\section{Evaluation}
\label{sec:eval}
\ga{Page budget: 2 pages \\}
\ga{Plan: we should discuss the robot aggregation scenario, specifying the state space, the action space, and the reward function.
I dunno if it makes sense to discuss the variants (i.e., fixed area, two areas and moving area). We will see.
}
\subsection{Scenario}
